On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-01 model.tex
regular update

\section{Single fluid description of well-coupled dusty gas}\label{setup} 
In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
\cite{laibe14} give a single fluid description of the mixture:  
(their Eq. 84 --- 85 and Eq. 87):

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla
  P, \label{momeq}\\ 
  &\frac{D\epsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\epsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \epsilon \equiv \frac{\rhod}{\rho}  = \frac{\tepsilon}{1+\tepsilon} 
\end{align}
is the dust-fraction and $\tepsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. The external potential $\Phi$ is
that for a star of mass $M_*$ at the origin 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant. The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

\subsection{Locally isothermal equation of state for the gas} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \epsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time.  
%For analytic discussion we consider arbitrary functions of
%$c_s(r,z)$. 
In numerical calculations we will consider vertically  
isothermal disks with 
\begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature.

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\epsilon$ in favor of $P$, 
\begin{align*}
  \epsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.

@@

For a given distribution of the dust-fraction $\epsilon$ (or
dust-to-gas ratio $\tepsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\epsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\epsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} 
except in the limit of perfectly coupled dust ($\tstop\to0$).

@@

%Eq. \ref{steady_dust} makes it difficult to obtain equilibrium
%solutions explicitly. However, if the diffusive process is slow and
%can be neglected, then one may just solve
%Eq. \ref{steady_momr}---\ref{steady_momz} with a prescribed (initial)
%distribution of the dust fraction $\epsilon(r,z)$.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
To obtain an actual disk structure for numerical computations, we
assume a Gaussian profile in the dust-to-gas ratio,   
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0(r)
  \exp{\left[-\frac{z^2}{2\widetilde{H}^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\tepsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

In practice, we consider $\tepsilon_0 \ll 1$ and settled dust such that
$\Htilde<\Hgas$. Then $\rhog(r,z)$ is effectively Gaussian, as in the
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \tepsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

\subsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\tepsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\epsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\epsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right),
\end{align}
where
\begin{align}
   s \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\epsilon)\right]} 
\end{align}
is a pseudo-entropy for the mixture.

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{
  \frac{\p\tepsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\tepsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\tepsilon}{\p z} + \left(1+\tepsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

We can compare these sources by
evaluating them, to order-of-magnitude, using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\tepsilon\ll1$. This gives

\begin{align}\label{vshear_split}
  \mathcal{V} =  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
  \tepsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(\tepsilon + \delta^2\right)},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio can be at
most comparable to that due to the radial temperature gradient; and
this only occurs for thin dust layers with $\delta^2\ll \tepsilon$ (so
that $\mathcal{V}\sim 1/|q|$). Otherwise, vertical shear is
predominantly due to $\p_rc_s^2$.

@@

\subsection{Strictly isothermal gas perfectly coupled to dust}   
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\epsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the Solberg-Hoiland criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\epsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\epsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \epsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align}
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency, and we assume $\kappa^2>0$. 
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are equivalent to the
Solberg-Hoiland criteria for adiabatic gas if one writes the 
entropy $s=\ln{\left[c_s^2\left(1 - \epsilon\right)\right]}$.

%These criteria are applicable to any adiabatic fluid. 
%For an isothermal mixture, we may take $s=\ln{(1-\epsilon)}$.   
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\epsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
%The stability conditions are thus
%In accretion disks, $\rhog$, $\rhod$ and $\epsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\epsilon$, and the dust-to-gas ratio $\tepsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

%We insert the vertical shear profile,
%Eq. \ref{vshear}, into Eq. \ref{dusty_solberg} and write the result 
%in terms of $\tepsilon$, to find

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the vertically-Gaussian
distributions for the gas density and the dust-to-gas ratio as 
described in \S\ref{eqm}. 
%However, here we permit 
%$\tepsilon_0=\tepsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
%radius. 
We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$. In terms of the dust-to-gas
ratio $\tepsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\tepsilon}{\left(1+
    \tepsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + 
  \frac{z^2}{\widetilde{H}^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\widetilde{H}^2}{H_\mathrm{g}^2}\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g},\tepsilon,\ll 1$ and $
\widetilde{H}/H_\mathrm{g}\lesssim 1$. Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

>>>>

\section{Single fluid description of dusty gas}\label{setup} 
In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
\cite{laibe14} give a single fluid description of the mixture:  
(their Eq. 84 --- 85 and Eq. 87):

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla
  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. The external potential $\Phi$ is
that for a star of mass $M_*$ at the origin 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant. The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

\subsection{Locally isothermal equation of state for the gas} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In numerical calculations we will consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature.

If we define the effective sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
    \tepsilon\right)}$, then the equation of state has the standard
form $P=c_{s,\mathrm{eff}}^2\rho$. Since the dust-fraction $\tepsilon$
typically decrease away from the midplane, we expect vertically
isothermal dusty disks to in fact behave as if it
had a temperture increasing away from $z=0$.

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.

@@

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} 
except in the limit of perfectly coupled dust ($\tstop\to0$).

@@

%Eq. \ref{steady_dust} makes it difficult to obtain equilibrium
%solutions explicitly. However, if the diffusive process is slow and
%can be neglected, then one may just solve
%Eq. \ref{steady_momr}---\ref{steady_momz} with a prescribed (initial)
%distribution of the dust fraction $\tepsilon(r,z)$.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
To obtain an actual disk structure for numerical computations, we
assume a Gaussian profile in the dust-to-gas ratio,   
\begin{align}\label{dust_gauss}
  \epsilon(r,z) = \epsilon_0(r)
  \exp{\left[-\frac{z^2}{2\Htilde^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\epsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \epsilon_0(r) = \epsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\epsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\epsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

In practice, we consider $\epsilon_0 \ll 1$ and settled dust such that
$\Htilde<\Hgas$. Then $\rhog(r,z)$ is effectively Gaussian, as in the
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \epsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

\subsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right),
\end{align}
where
\begin{align}
   s \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]} 
\end{align}
is a pseudo-entropy for the mixture.

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

We can compare these sources by
evaluating them, to order-of-magnitude, using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\epsilon\ll1$. This gives

\begin{align}\label{vshear_split}
  \mathcal{V} =  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
  \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio can be at
most comparable to that due to the radial temperature gradient; and
this only occurs for thin dust layers with $\delta^2\ll \epsilon$ (so
that $\mathcal{V}\sim 1/|q|$). Otherwise, vertical shear is
predominantly due to $\p_rc_s^2$.

\subsection{Vertical buoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p s}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. We find 
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
Thus in PPDs where $\epsilon_0 = O(10^{-1}$---$10^{-2})$, 
vertical buoyancy only has a modestly stabilizing effect.

%\begin{align}
 % N_z^2 
%\end{align}

@@

\subsection{Strictly isothermal gas perfectly coupled to dust}   
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the Solberg-Hoiland criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align}
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency, and we assume $\kappa^2>0$. 
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are equivalent to the
Solberg-Hoiland criteria for adiabatic gas if one writes the 
entropy $s=\ln{\left[c_s^2\left(1 - \tepsilon\right)\right]}$.

%These criteria are applicable to any adiabatic fluid. 
%For an isothermal mixture, we may take $s=\ln{(1-\tepsilon)}$.   
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\tepsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
%The stability conditions are thus
%In accretion disks, $\rhog$, $\rhod$ and $\tepsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\tepsilon$, and the dust-to-gas ratio $\epsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

%We insert the vertical shear profile,
%Eq. \ref{vshear}, into Eq. \ref{dusty_solberg} and write the result 
%in terms of $\epsilon$, to find

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the vertically-Gaussian
distributions for the gas density and the dust-to-gas ratio as 
described in \S\ref{eqm}. 
%However, here we permit 
%$\epsilon_0=\epsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
%radius. 
We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\epsilon}{\left(1+
    \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g},\epsilon,\ll 1$ and $
\Htilde/H_\mathrm{g}\lesssim 1$. Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.
